Building Markov State Models for Periodically Driven Non-Equilibrium Systems

### 2014-09-22 writedown/period.noneq.msm/paper.tex
right data of matrix
revise the conclusion accordingly

We define the one-step transition matrix $\hat{\vect P}$ for the milestone process by
\begin{equation}\label{hatP_prob}
\hat{\vect P}_{jk}=\prob_\mu\Big(\hat{X}_{m+1}=k\,\vert\, \hat{X}_m=j\Big).
\end{equation}
Then Following
Ref.~\cite{djurdjevac2010markov}, Thm.~3.1, 
$\hat{\vect P}$ can be computed by matrix multiplication using the committors:
  \begin{align}
    \label{eq:msm-tmatrix-00}
    \begin{split}      
    \hat{\vect P}_{jk}
    = &
    \frac{1}{\hat{\mu}(j)}
    \langle (\vect P^b - \id) q^\bwd_j,q^\fwd_k \rangle_\mu,\qquad j\not= k, \\    %\label{eq:msm-tmatrix-01}
    \hat{\vect P}_{jj}
    =&
    1-\sum_{k\not=j} \hat{\vect P}_{jk}
    \end{split}
  \end{align}
where the inner product is defined by
$\langle u,v \rangle_\mu=\sum_{i\in S} u(i) v(i) \mu(i)$. 
In general the milestone process need not be a Markov process. The results in \cite{A19-31,sarich2014utilizing} show, however, that it is an approximate Markov process as long as the core sets are proper metastable sets, i.e., if the typical timecale on which $(X_m)$ leaves $C$ is much smaller than the typical expected hitting times between the core sets.
Thus, by taking $\hat{P}$ as our MSM transition matrix, we introduce an additional modelling error that is the smaller the more metastable the core sets are.
With this MSM transition matrix, we can define the MSM kinetics:
If we start from some initial probability $\hat{p}_j(0)$ of being in state $j$ at time $t=0$ then its evolution $\hat{p}_j(t)$ in time is discrete in multiples of period $T$ and given by
\begin{equation}\label{eq:num-29}
\hat{p}_j(T)=\sum_k \hat{p}_k(0)\hat{\vect P}_{kj}.
\end{equation}
In our approach $\hat p_j(mT)$ is a good approximation of $\prob(\hat{X}_m=j)$ (for appropriately chosen core sets).

@@

Following the process described in Sec.~\ref{sec:build-msm}, we are able to build a 
three state MSM for the externally driven alanine dipeptide system,
where the quasi-stationary probability distribution $\mu$, the three core sets, and the forward and backward committors are
estimated as described above, and the MSM transition matrices $\hat P$, $\hat{M}$, and $\hat{T}$ are then evaluated using Eq.~\eqref{eq:msm-tmatrix-00}, \eqref{eq:msm-ht-1} and \eqref{eq:msm-hm-1}, correspondingly.
Alternatively, the MSM  transition matrix $\hat P$ is calculated directly from the NEMD trajectories using Eq.~\eqref{hatP_prob} (very good agreement with the one computed from the committors).
The leading eigenvalues of $P=\myphi^{\top}(T)$ are compared with those of $\hat P$ and $\hat{M}^{-1}\hat{T}$ in
Tab.~\ref{tab:tmp1}.
Without surprising, the two approaches for MSM building are consitent, \recheck{and the latter is slightly better}.
The MSM is able to accurately reproduce
the largest non-trivial eigenvalue, which means a precise reproduction
of the longest non-trivial implied time-scale. The accuracy of the second non-trivial
time-scale is not as good as the first, but is still acceptable.
The reason
for the lower accuracy may also be that
the corresponding time scale is 26.5~ps (calculated by $-T/\log(\lambda_2)$),
which is NOT significantly longer than the temporal resolution given by the period $T=10$~ps of the external driving force.
The difference between the eigenvalues of $\hat P$ and $\hat{M}^{-1}\hat T$
  can be taken as an indication for the non-Markovianity of the milestone process $\hat X_m$. This
  non-Markovianity seems to have stronger influence on the second non-trivial time-scale in comparison to the first one; this may be caused by shorter decorrelation timescales due to weaker metastability of the core sets involved.
Numerically the two MSM transition matrices are:
\[
\hat{M}^{-1}\hat{T}=\left(\begin{array}{ccc}
   0.860 &   0.134 &   0.007 \\
    0.190  &  0.775  &  0.034\\
    0.020   & 0.065 &   0.915
\end{array}
\right),\qquad \hat{P}=\left(\begin{array}{ccc}
   0.882 &   0.110 &   0.008 \\
    0.158  &  0.815  &  0.028\\
    0.023   & 0.055 &   0.922
\end{array}
\right),
\]
from which we see that  the left-handed alpha-helix conformations of the peptide exhibits the strongest metastability.

@@

In this paper we developped methods of building MSM for a periodically driven
non-equilibrium system. By a numerical example
of the alanine dipeptide under periodically oscillatory EF,
the methods are shown  to be
valid and be able to capture the long-time behavior of the original
non-equilibrium dynamics. We have provided and verified two equivalent
pathways of building MSM, as is shown by the two branches in
Fig.~\ref{fig:flowchart}.

The the key steps of the first pathway are
summarized as following:
\begin{enumerate}
\item By a fine discretization of the molecular configuration space,
  the original NEMD dynamics is discretized into a (in principle)
  time-continuous \recheck{Markov} jump process with a time-dependent generator $L(t)$.
  If the discretization is good enough, then the discretized
  dynamics is expected to reproduce the original dynamics.
  In principle most of the discretiztion methods for the
  equilibirum case can be generalized to  non-equilibrium without
  substantial difficulty, however,
  we have not discussed which discretiztion method is the best choice,
  since it is out of the scope of the present work.
  \recheck{It should be noticed that
    we assume the jump process is Markovian, althogh in
    general it is not true. However, if (1) the system present time-scale
    seperation; (2) the spatial discretiztion is fine enough; (3)
    we approximately consider a discretizted time
    (can be very fine, and in computer the molecular configurations are
    saved at discretizted time steps)
    by lag-time $\tau$, so that within the lag-time
    the fast dynamics of the system is fully relaxed; then
    the Markovianity is a good assumption for the discretizted dynamics.
  }
  We show that, by numerical example, a uniform $20\times20$ discretiztion 
  over the dihedral angles $\phi$--$\psi$ space is a good choice for
  alanine dipeptide.
  In the tests, the state independency in computing $L(t)$
  is used as an indicator for
  the quality of the discretiztion.
\item By the Floquet thoerem, the time-inhomogeneous process generated
  by $L(t)$ is temporally discretized in to a time-homogeneous,
  but not necessarily reversible, jump process that has a transition
  matrix denoted by $P$. This temporal discretiztion only  considers
  the state of the system at integral periods, therefore all information
  within the period is not preserved. This is, however, not a serious
  problem whenever the long-time hehaviors of the system is of interst, and
  their corresponding time-scales are significantly longer than one period.
  The advantage of this discretiztion
  is also obvious: All well-developed theories and computational tools
  for time-homogeneous Markov processes
  can be directly appled. In the
  alanine dipeptide system, we compare
  the the quasi-stationary probability,
  first mean hitting time, and the forward and backward committors
  of the discretized dynamics with those computed directly from the
  NEMD simulation. An almost perfect consistency 
  indicates a good approximation quality of the discretized dynamics.
\item From the disretized dynamics, building the MSM is straightforward
  by using Eq.~\eqref{eq:msm-tmatrix-00} and a set of coresets.
  Numerical results shows that a 3-state MSM can reporduce the leading
  non-trivial eigenvalue with very good accuracy, while the second
  non-trivial eigenvalue with acceptable accuracy. The lower accuracy
  regarding the second non-trivial eigenvalue may be due to the fact that
  the second slowest time-scale is not significantly longer than the
  period. Using this three state MSM, we can also reproduce, with
  almost perfect accuracy, the 
  time-evolution of the metastable state
  pobabilities from equilibirum (no external driven)
  to the non-equilibrium quasi-stationary state.
\end{enumerate}

The right branch in Fig.~\ref{fig:flowchart} presents a
equivalent, and seemingly
much simpler alternative to the left branch, provided the
coresets are known. Neverthless, in pactice, identifying
the coresets is not a trivial task, especially for the
non-equilibrium molecular systems.
Therefore, a good discretization is  usually a must for computing the coreset.
For example, in the present work, we computed the coresets by
finding the roughly constant levels of the hitting time distribution.
This method only works for the discretized dynamics $P$, and its continuous
counterpart is still an open question.

>>>>

We define the one-step transition matrix $\hat{\vect P}$ for the milestone process by
\begin{equation}\label{hatP_prob}
\hat{\vect P}_{jk}=\prob_\mu\Big(\hat{X}_{m+1}=k\,\vert\, \hat{X}_m=j\Big).
\end{equation}
Then Following
Ref.~\cite{djurdjevac2010markov}, Thm.~3.1, 
$\hat{\vect P}$ can be computed by matrix multiplication using the committors:
  \begin{align}
    \label{eq:msm-tmatrix-00}
    \begin{split}      
    \hat{\vect P}_{jk}
    = &
    \frac{1}{\hat{\mu}_j}
    \langle (\vect P^b - \id) q^\bwd_j,q^\fwd_k \rangle_\mu,\qquad j\not= k, \\    %\label{eq:msm-tmatrix-01}
    \hat{\vect P}_{jj}
    =&
    1-\sum_{k\not=j} \hat{\vect P}_{jk}
    \end{split}
  \end{align}
where the inner product is defined by
$\langle u,v \rangle_\mu=\sum_{i\in S} u(i) v(i) \mu(i)$. 
In general the milestone process need not be a Markov process. The results in \cite{A19-31,sarich2014utilizing} show, however, that it is an approximate Markov process as long as the core sets are proper metastable sets, i.e., if the typical timecale on which $(X_m)$ leaves $C$ is much smaller than the typical expected hitting times between the core sets.
Thus, by taking $\hat{P}$ as our MSM transition matrix, we introduce an additional modelling error that is the smaller the more metastable the core sets are.
With this MSM transition matrix, we can define the MSM kinetics:
If we start from some initial probability $\hat{p}_j(0)$ of being in state $j$ at time $t=0$ then its evolution $\hat{p}_j(t)$ in time is discrete in multiples of period $T$ and given by
\begin{equation}\label{eq:num-29}
\hat{p}_j(T)=\sum_k \hat{p}_k(0)\hat{\vect P}_{kj}.
\end{equation}
In our approach $\hat p_j(mT)$ is a good approximation of $\prob(\hat{X}_m=j)$ (for appropriately chosen core sets).

@@

Following the process described in Sec.~\ref{sec:build-msm}, we are able to build a 
three state MSM for the externally driven alanine dipeptide system,
where the quasi-stationary probability distribution $\mu$, the three core sets, and the forward and backward committors are
estimated as described above, and the MSM transition matrices $\hat P$, $\hat{M}$, and $\hat{T}$ are then evaluated using Eq.~\eqref{eq:msm-tmatrix-00}, \eqref{eq:msm-ht-1} and \eqref{eq:msm-hm-1}, correspondingly.
Alternatively, the MSM  transition matrix $\hat P$ is calculated directly from the NEMD trajectories using Eq.~\eqref{hatP_prob} (very good agreement with the one computed from the committors).
The leading eigenvalues of $P=\myphi^{\top}(T)$ are compared with those of $\hat P$ and $\hat{M}^{-1}\hat{T}$ in
Tab.~\ref{tab:tmp1}.
Without surprising, the two approaches for MSM building are consitent.
The MSM is able to accurately reproduce
the largest non-trivial eigenvalue, which means a precise reproduction
of the longest non-trivial implied time-scale. The accuracy of the second non-trivial
time-scale is not as good as the first, but is still acceptable.
The reason
for the lower accuracy may also be that
the corresponding time scale is 26.5~ps (calculated by $-T/\log(\lambda_2)$),
which is NOT significantly longer than the temporal resolution given by the period $T=10$~ps of the external driving force.
The difference between the eigenvalues of $\hat P$ and $\hat{M}^{-1}\hat T$
  can be taken as an indication for the non-Markovianity of the milestone process $\hat X_m$. This
  non-Markovianity seems to have stronger influence on the second non-trivial time-scale in comparison to the first one; this may be caused by shorter decorrelation timescales due to weaker metastability of the core sets involved.
Numerically the two MSM transition matrices are:
\[
\hat{M}^{-1}\hat{T}=\left(\begin{array}{ccc}
   0.860 &   0.133 &   0.008 \\
    0.192  &  0.775  &  0.033\\
    0.019   & 0.066 &   0.915
\end{array}
\right),\qquad \hat{P}=\left(\begin{array}{ccc}
   0.882 &   0.110 &   0.008 \\
    0.158  &  0.815  &  0.028\\
    0.023   & 0.055 &   0.922
\end{array}
\right),
\]
from which we see that  the left-handed alpha-helix conformations of the peptide exhibits the strongest metastability.

@@

In this paper we developped methods of building MSM for a periodically driven
non-equilibrium system. By a numerical example
of the alanine dipeptide under periodically oscillatory EF,
the methods are shown  to be
valid and be able to capture the long-time behavior of the original
non-equilibrium dynamics.

We firstly provide effective methods for discretizing the original NEMD trajectories
both temporally and spacially, and the end-product is a time-homogeneous
(not necessarily reversible) Markov jump process.
There are two equivalent approaches discussed in the paper:
one can either do a two-steps discretization,
i.e.~firstly spacial and then temperoal,
or a single step  involes both discretizations simulteneously.
They are exhibited by the left-most and middle branches in Fig.~\ref{fig:flowchart}.

By a fine discretization of the molecular configuration space, the
original NEMD dynamics is discretized into a (in principle)
continuous and time-inhomogeneous {Markov} jump process.
In principle most of the discretiztion methods for the equilibirum
case can be generalized to non-equilibrium without substantial
difficulty, however, we have not compare different discretiztion methods
regarding the accuracy, since it is out of the scope of the present work.
We show that, by numerical example, a uniform $20\times20$
discretiztion over the dihedral angles $\phi$--$\psi$ space is a good
choice for alanine dipeptide.  In the tests, the state independency in
computing the generator is used as an indicator for the quality of the
discretiztion.

The Floquet thoerem discretize a periodic process by taking only
integral periods, so that
the time-inhomogeneous process
is discretized in to a time-homogeneous, but not
necessarily reversible process.
The advantage of this discretiztion is  obvious: All
well-developed theories and computational tools for time-homogeneous
Markov processes can be directly appled. In the alanine dipeptide
system, we compare the the quasi-stationary probability, first mean
hitting time, and the forward and backward committors of the
discretized dynamics with those computed directly from the NEMD
simulation. An almost perfect consistency indicates a good
approximation quality of the discretized dynamics.

Although the end-product of the two-step and one-step discretiztions are equivalent,
it is clear that the one-step discretiztion does not preserve any information
within one period, because only the states
at integral periods are considered. This is not a serious problem
whenever the long-time hehaviors of the system is of interst, and
their corresponding time-scales are significantly longer than one
period. However, if the time-scale of any interested behavior
is comparable to, or even shorter than the period, the two-step
discretiztion is prefered, so that one can recover the dynamics within
one period no matter when it is necessary. If the longest non-trivial
time-scale is comparable to one period, thenn the Floquet theorem
is not preferable, and the non-equilibrium dynamics may be well
described by the quasi-equilibrium approximation.

Building the MSM from the time-homogeneous disretized dynamics is straightforward by
using Eq.~\eqref{eq:msm-tmatrix-00} and a set of coresets that are
derived from the discretized dynamics.  Numerical
results shows that a 3-state MSM can reporduce the leading non-trivial
eigenvalue with very good accuracy, while the second non-trivial
eigenvalue with acceptable accuracy. The lower accuracy regarding the
second non-trivial eigenvalue may be due to the fact that the second
slowest time-scale is not significantly longer than the period. Using
this three state MSM, we can also reproduce, with almost perfect
accuracy, the time-evolution of the metastable state pobabilities from
equilibirum (no external driven) to the non-equilibrium
quasi-stationary state.

The right-most branch in Fig.~\ref{fig:flowchart} presents an
equivalent, and seemingly
much simpler alternative to the middle branch, i.e. building
the MSM directly from the NEMD trajectories.
In practice, however, this method may not be applicable, because
it requires known coresets, and the coresets identification 
is usually not a trivial task, especially for the
non-equilibrium molecular systems.
This task is substantially
simplified when an accurate time-homogeneous discretiztion 
to the original time-inhomogeneous NEMD is accessible.
In the present work, we computed the coresets by
finding the roughly constant levels of the hitting time distribution 
of the discretized dynamics.
It should be noticed that the continuous
counterpart of the current coresets identification method is still an open question.
% This method only works for the discretized dynamics. and its
% Therefore, a good discretization is  usually a must for computing the coreset.
